Assessment and Optimization of Nowcast Measures for Trend Detection

### 2023-12-04 paper.tex
New outline of paper

\documentclass{article}

@@

%%% --- Figures and Graphics
\usepackage{graphicx}
\usepackage{subcaption}

%%% --- Colors
\usepackage{color}
\usepackage[dvpinames]{xcolor}

%%% --- Tabulars
\usepackage{booktabs}

@@

%%% --- Commenting, Todonotes
\usepackage{comment, todonotes}
\newcommand{\ar}{$\rightarrow$}
\presetkeys{todonotes}{backgroundcolor=yellow!30, bordercolor=yellow!50, linecolor=yellow!50, figwidth=\textwidth}{}
\newcommand{\hl}[1]{\textcolor{Aquamarine}{#1}}

@@

\newcommand{\lag}[1][l]{\Delta_{#1}}
\DeclarePairedDelimiter{\abs}\lvert\rvert

@@

\title{Trending in Nowcasting}
\author{Oliver Grothe, Bolin Liu, Jonas Rieger}

@@

%\begin{abstract}
%Your abstract.
%\end{abstract}

\input{sections/01-introduction.tex}

\section{Notation}
\begin{itemize}
  \item Sei $T = \{1, 2, \dots\}$ die Zeitindexmenge, wobei für jeden Zeitpunkt $t \in T$ eine Realisierung vorliegt (die ggf. im Nachhinein veröffentlicht wird).
  \item Sei $(Y_t)_{t \in T}$ die Zeitreihe der Realisierungen, dabei steht $t$ für den Zeitpunkt, auf den sich der Wert bezieht, nicht den Veröffentlichungszeitpunkt.
  \item Sei $K = \{1, 2, \dots\}$ die Menge der Nowcaster.
	\begin{itemize}
	  \item Sei $X_{t \lvert \tau}^k (k \in K, t \in T, \tau \in T_t^k)$ der Nowcast von $k$ bezüglich des Zeitpunktes $t$, der am Zeitpunkt $\tau$ \textbf{veröffentlicht} wird (oder: berechnet / dessen Informationen sich auf Zeitpunkt $\tau$ beziehen).
	  \item Sei $g: \mathbb{R}^{\lvert T_t^k \lvert} \rightarrow \mathbb{R}$ eine Aggregationsfunktion, die alle Nowcasts bezüglich eines Zeitpunkts zu einem Nowcast zusammenfasst ($X_t^k \coloneqq g((X_{t \lvert \tau}^k)_{\tau \in T_t^k})$). Falls jeweils nur ein Nowcast veröffentlich wird, wähle $g(x) = x$.
	  \item Sei $(X_t^k)_{t \in T}$ die Zeitreihe der aggregierten Nowcasts.
	\end{itemize}
  \item Sei $\lag[l]$ der lag-$l$-Operator, der für eine Zeitreihe $(Z_t)_{t \in T}$ definiert wird durch 
		\begin{equation}
			\lag[l]Z_t = Z_{t+l} - Z_t \quad (t \in T)
		\end{equation} 
  \item Der Anteil an konkordanten Punkten für Realisierungen $y = (y_1, y_2, \dots, y_n)$ und $(x_1, x_2, \dots, x_n)$ bezüglich des Lag $l$ ergibt sich dann durch
	\begin{equation}
  		k (x, y; l, \epsilon) = \frac{\sum_{t}^{n-l} k^s (\lag y_t, \lag x_t; l, \epsilon)}{\sum_{t}^{n-l} k^\epsilon (\lag y_t, \lag x_t)}
	\end{equation}
	wobei $k^s$ die Indikatorfunktion für die Konkordanz zweier Werte außerhalb der exclusion area ist \todo{stimmt Konkordanz hier?}
	\begin{equation}
  		k^s (x, y; \epsilon) \coloneqq \begin{cases}
  			1 &, \text{falls}\ xy > 0\ \text{und}\ k^\epsilon(x, y; \epsilon) = 1\\
  			0 &, \text{sonst}
  		\end{cases}
	\end{equation}
	und $k^\epsilon$ die Indikatorfunktion für Punkte außerhalb der exclusion area ist
	\begin{equation}
  		k^\epsilon (x, y; \epsilon) \coloneqq \begin{cases}
  			1 &, \text{falls} \ \abs{x} > \epsilon \text{oder} \ \abs{y} > \epsilon \\
  			0 &, \text{sonst}
  		\end{cases}
	\end{equation}
\end{itemize}

\section{Statistische Modellierung}

\begin{itemize}
  \item Sei $l$ der zeitliche Abstand, mit dem Realisationen verfügbar werden
  \item $Y_{t + l} = Y_t + \lag Y_t$
  \item Die Nowcast schätzen (fehlerbehaftet) $\lag Y_t$ mithilfe von Daten, die zum Zeitpunkt $\tau \in T_t^k$ verfügbar sind, und dem neuesten, bekannten Wert $Y_t$:
	\begin{equation}
  		X_{t+l \lvert \tau}^k = Y_t + (\widehat{\lag Y_t})_{t \lvert \tau \in T_t^k}^k,
	\end{equation}
	\todo{Stimmt das so?}
	\todo{Ist festes $l$ zu starke Annahme?}
	und damit 
		\begin{equation}
  			\lag Y_t = (\widehat{\lag Y_t})_{t \lvert \tau \in T_t^k}^k + \varepsilon.
		\end{equation}

\end{itemize}

\section{Trending}

\begin{itemize}
  \item \enquote{Schwaches} Trending für lag $l$:
   \begin{equation}
  	P(\lag Y_t \widehat{(\lag Y_t)^k} > 0 \:\lvert\: \abs{\lag Y_t} > \epsilon) > P(\lag Y_t \widehat{(\lag Y_t)^k} < 0 \:\lvert\: \abs{\lag Y_t} > \epsilon)
   \end{equation}

\end{itemize}

%\printbibliography

>>>>

\documentclass[oneside]{article}

@@

%%% --- Figures and Graphics
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{tikz}

%%% --- Colors
\usepackage[]{xcolor}

%%% --- Tabulars
\usepackage{booktabs}
\usepackage{multirow} % cells over several rows
\usepackage{tabularx}

@@

%%% --- Commenting, Todonotes
\usepackage{comment, todonotes, soul}
\newcommand{\ar}{$\rightarrow$}
\presetkeys{todonotes}{backgroundcolor=yellow!30, bordercolor=yellow!50, linecolor=yellow!50, figwidth=\textwidth}{}
%\newcommand{\hl}[1]{\textcolor{Aquamarine}{#1}}

@@

\newcommand{\lag}[1][l]{\Delta_{#1}}
\DeclarePairedDelimiter{\abs}\lvert\rvert
\DeclareMathOperator{\sign}{sign}
\newcommand{\tmax}{\bar{t}}
\newcommand{\ind}[1]{\mathbbm{1}\{#1\}}
\newcommand{\ydiff}{\tilde{y}}
\newcommand{\ydifft}{\tilde{y}^*}
\newcommand{\xdiff}{\tilde{x}}
\newcommand{\xdifft}{\tilde{x}^*}
\newcommand{\Ydiff}{\tilde{Y}}
\newcommand{\Xdiff}{\tilde{X}}
\newcommand{\Prob}[1]{P(#1)}
\newcommand{\mprob}{\tilde{m}}

@@

%%% --- ToDo List
\usepackage{enumitem}
\newlist{todolist}{itemize}{2}
\setlist[todolist]{label=$\square$}
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\done}{\rlap{$\square$}{\raisebox{2pt}{\large\hspace{1pt}\cmark}}%
\hspace{-2.5pt}}
\newcommand{\wontfix}{\rlap{$\square$}{\large\hspace{1pt}\xmark}}

\title{Assessment and Optimization of Nowcast Measures for Trend Detection}
\author{Oliver Grothe, Bolin Liu, Jonas Rieger}

@@

\begin{abstract}
Existing and new measures for the capability of capturing the trend of nowcasts are presented, evaluated and compared on synthetic and real-world data.
\end{abstract}

\section{Introduction}

\begin{itemize}
	\item Trending is an important feature of nowcasts to be identify changes in the situation before the actual quantity can be measured
	\item Examples: Covid, BIP, Medicine, ... 
\end{itemize}

\section{Trend measures for nowcasts}

Notation
\begin{itemize}
  \item Let $y_1, \ldots, y_T$ be the values of the nowcasted quantity.
  \item Let $x_t^k$ be the $k$-th nowcast for time $t = 1, \ldots, T$ (possibly after some aggregation if several raw nowcasts apply).
  \item Let $\ydiff_t = y_t - y_{t-1}$ be the change occurred between time steps $t-1$ and $t$ and $\xdiff_t^k = x_t - y_{t-1}$ the change nowcaster $k$ issued between the time steps. 
  \item Then, $\xdiff_t^k$ can be seen as point forecast of $\ydiff_t$. 
\end{itemize}

\subsection{What is the trend of a nowcast?}

Trend: Issuing the right \enquote{direction} of change between two time steps

\begin{itemize}
  \item Translate into mathematical notation
  \item Illustrative Example: Two nowcasts of quantity of interest, where one jumps up and down and one points in right direction with the same rmse
\end{itemize}

\subsection{Measures}

%\begin{table}
%	\centering
%	\begin{tabular}{l l c c}
%	\toprule
%	& & \multicolumn{2}{c}{Observed trend} \\
%	& & $\ydiff_t > 0$ & $\ydiff_t < 0$ \\	\cline{3-4}
%	\multirow{2}{*}{Forecasted trend} & $\xdiff_t > 0$ & hit & false alarm\\
%	& $\xdiff_t < 0$ & miss & correct negative \\
%	\bottomrule
%\end{tabular}
%\end{table}

Non-probabilistic measures:

\begin{itemize}
  \item Accuracy: \begin{equation}
  	m_{\text{acc}} = \frac{\sum_{t=2}^T \ind{\ydiff_t > 0, \xdiff_t > 0} + \sum_{t=2}^T \ind{\ydiff_t < 0, \xdiff_t < 0}}{T-1}
\end{equation}
	\item Capability of detecting a positive trend (probability of detection)
	\begin{equation}
  		m_{\text{py}} = \frac{\sum_{t=2}^T \ind{\ydiff_t > 0, \xdiff_t > 0}}{\sum_{t=2}^T \ind{\ydiff_t > 0}}
	\end{equation}
	\item False alarm rate \hl{Name does not fit} \begin{equation}
  m_{\text{far}} = \frac{\sum_{t=2}^T \ind{\ydiff_t < 0, \xdiff_t > 0}}{\sum_{t=2}^T \ind{\xdiff_t > 0}}
\end{equation}
maybe use \enquote{positive} measure instead, i.e.,
\begin{equation}
  m_{\text{px}} = \frac{\sum_{t=2}^T \ind{\ydiff_t > 0, \xdiff_t > 0}}{\sum_{t=2}^T \ind{\xdiff_t > 0}}  
\end{equation}
	\item same measures for negative direction:
	\begin{align}
		m_{\text{ny}} &=  \frac{\sum_{t=2}^T \ind{\ydiff_t < 0, \xdiff_t < 0}}{\sum_{t=2}^T \ind{\ydiff_t < 0}} \\
		m_{\text{nx}} &=  \frac{\sum_{t=2}^T \ind{\ydiff_t < 0, \xdiff_t < 0}}{\sum_{t=2}^T \ind{\xdiff_t < 0}}
	\end{align}
\end{itemize}

Resulting measures:
\begin{itemize}
  \item Scoring rules for probabilities

\begin{itemize}
  \item Quadratic/Brier Score for $k$ possible outcomes: \begin{equation}
  S((p_1, \dots, p_k), i) = 2 p_i - \sum_{l=1}^k p_i^2 - 1;
\end{equation}
for dichotomous outcomes ($p_1 = \hat{P}(\ydiff_t > 0)$: probability for $\ydiff_t > 0$):
\begin{equation}
  S(p_1, \ydiff_t) = p_1 \ind{\ydiff_t > 0} + (1 - p_1) (1 - \ind{\ydiff_t > 0}) + p_1^2 + (1-p_1)^2 - 1
\end{equation}
\end{itemize}
\item Probabilistic versions of non-probabilistic measures:
\begin{itemize}
  \item Assume an additive error decomposition for both observation and nowcast to account for non-systematic and short-term (i.e., intra-day) effects:
  	\begin{align}
  		\ydiff_t &= \ydifft + \varepsilon_t^y \\
  		\xdiff_t &= \xdifft + \varepsilon_t^x
	\end{align}
\item  Replace indicators in the above formulations by the corresponding probabilities:
\begin{align}
		\mprob_{\text{acc}} &= \frac{\sum_{t=2}^T \Prob{ \ydifft_t > 0, \xdifft_t > 0} + \sum_{t=2}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{T-1}  \\
   \mprob_{\text{py}} &= \frac{\sum_{t=2}^T \Prob{\ydifft_t > 0, \xdifft_t > 0}}{\sum_{t=2}^T \Prob{\ydifft_t > 0}} = \sum_{t=2}^T \frac{\Prob{\ydifft_t > 0, \xdifft_t > 0}}{\Prob{\ydifft_t > 0}} = \sum_{t=2}^T \Prob{\xdifft_t > 0 | \ydifft_t > 0} \\
    \mprob_{\text{px}} &= \frac{\sum_{t=2}^T \Prob{\ydifft_t > 0, \xdifft_t > 0}}{\sum_{t=2}^T \Prob{\xdifft_t > 0}} = \sum_{t=2}^T \Prob{\ydifft_t > 0 | \xdifft_t > 0} \\
    \mprob_{\text{ny}} &= \frac{\sum_{t=2}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{\sum_{t=2}^T \Prob{\ydifft_t < 0}} = \sum_{t=2}^T \Prob{\xdifft_t < 0 | \ydifft_t < 0} \\
    \mprob_{\text{px}} &= \frac{\sum_{t=2}^T \Prob{\ydifft_t < 0, \xdifft_t < 0}}{\sum_{t=2}^T \Prob{\xdifft_t < 0}} = \sum_{t=2}^T \Prob{\ydifft_t < 0 | \xdifft_t < 0}
\end{align}
\item Simple error model:
  \begin{align}
	  \varepsilon_Y \sim N(0, \sigma_Y) \\
	  \varepsilon_X \sim N(0, \sigma_X)
  \end{align}
  yields, e.g.,
  	\begin{equation}
  		\mprob_{\text{acc}} = \frac{\sum_{t=2}^T  \big( 1 - \Phi_{\ydiff_t, \sigma_y}(0) - \Phi_{\xdiff_t, \sigma_x} (0) + 2 \Phi_{(\ydiff_t, \xdiff_t); (\sigma_y, \sigma_x)}( 0, 0) \big) }{T-1}, 
	\end{equation}
	where $\Phi$ denotes a (possibly multivariate) normal distribution
\end{itemize}
\end{itemize}

\subsection{Confidence intervals for the measures}

\section{Simulation studies}

\subsection{Different trending abilities for same rmse and mae} \label{sec:simulation_rmse_mae}

Let the quantity of interest 
\begin{equation}
  y_t = a \sin(2 \pi t / T) + \varepsilon_t^y, \ t = 1, \dots, T
\end{equation}
with $\varepsilon_t^y \stackrel{\text{iid}}{\sim} N(0, \sigma_y)$ and 
\begin{equation}
  \ydiff_t = y_t - y_{t-1}.
\end{equation}
The nowcasts are characterised by 
\begin{align}
	\xdiff_t^1 &= e^x_t \ydiff_t \\
	\xdiff_t^2 &= \begin{cases}
		e^x_t \ydiff_t &, \text{if}\ q^2_t = 0 \lor e^x_t > 0\\
		(| e^x_t | + 2) \ydiff_t &, \text{if}\ q^2_t = 1 \land e^x_t < 0
	\end{cases} \\
	\xdiff_t^3 &= \begin{cases}
		e^x_t \ydiff_t &, \text{if}\ (q^{3, \text{down}}_t = 0 \lor e^x_t > 0) \land \ydiff_t \geq 0\\
		(| e^x_t | + 2) \ydiff_t &, \text{if}\ (q^{3, \text{down}}_t = 1 \land e^x_t < 0) \land \ydiff_t \geq 0 \\
		e^x_t \ydiff_t &, \text{if}\ (q^{3, \text{up}}_t = 0 \lor e^x_t > 0) \land \ydiff_t < 0\\
		(| e^x_t | + 2) \ydiff_t &, \text{if}\ (q^{3, \text{up}}_t = 1 \land e^x_t < 0) \land \ydiff_t < 0
	\end{cases}
\end{align}
where $q^2_t \sim \text{Ber}(p_2)$ and $p_2$ denotes the probability that the direction of nowcast 2 is corrected and for nowcast 3 with $q^{3, \text{up}}_t \sim \text{Ber}(p_3^{\text{up}})$, $q^{3, \text{down}}_t \sim \text{Ber}(p_3^{\text{down}})$ the probabilities of direction correction depend on the direction of change.

\begin{figure}
  \centering
  \includegraphics{plots/simulation_same_rmse_mae/time_series.pdf}
  \caption{Realisation of Simulation~\ref{sec:simulation_rmse_mae} with $T = 100$, $\sigma_y=0.1$, $a = 10$, $p_2 = 0.8$, and $p_3 = (0.8, 0.4)$}
  \label{fig:simulation_rmse_mae_ts}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{plots/simulation_same_rmse_mae/4q_plot.pdf}
  \caption{4Q-Plot for Simulation~\ref{sec:simulation_rmse_mae} with $T = 100$, $\sigma_y=0.1$, $a = 10$, $p_2 = 0.8$, and $p_3 = (0.8, 0.4)$.}
  \label{fig:simulation_rmse_mae_4q}
\end{figure}

\hl{Add results}

\section{Application on disease and economic data}

\section{Conclusion}
